---
title: "Antidepressant effects of esketamine in adolescents with major depressive disorder and suicidal ideation: a case series"
subtitle: Reproducible analysis code
author:
 - Daniela Faria-Guimarães
 - Lucca S. Souza
 - Breno Souza-Marques
 - Flávia Vieira
 - Igor D. Bandeira
 - Samantha S. Silva
 - Raíza Alves-Pereira
 - Mariana Fontes
 - Rodrigo P. Mello
 - Gustavo C. Leal
 - Acioly L. T. Lacerda
 - Flávio Kapczinski
 - Taiane A. Cardoso
 - Aline S. Sampaio
 - Lucas C. Quarantini
output: 
    bookdown::pdf_document2:
      number_sections: true
      toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Note on reproducibility

This R Markdown file will exactly reproduce our data tidying, transformation, analysis and graphing processes in the R software and produce an illustrative PDF file. Code-only scripts are also available in the Open Science Framework repository for this study, at [https://osf.io/2ujqn/](https://osf.io/2ujqn/). The reproduction of this R Markdown file requires that the dataset available in the OSF repository is located in the same directory of this file, and that the user has the necessary dependencies to create a PDF using the `{bookdown}` package. Further information is available at the package website: [https://bookdown.org](https://bookdown.org). Please refer to the repository for a detailed description of each variable contained herein. All files in the repository, including this one, are licensed under [CC-BY-4.0](https://creativecommons.org/licenses/by/4.0/).

The `{renv}` package was used to help with reproducibility. The `renv.lock` file is available and should be imported to the same location on your computer as this script. For further information on how to use these files visit the package website: [https://rstudio.github.io/renv/articles/renv.html](https://rstudio.github.io/renv/articles/renv.html).

# Load the required packages


```{r, warning=F, message=F}
# Required Packages
library(tidyverse)
library(haven)
library(lsr)
library(rstatix)
library(boot)
library(kableExtra)
library(patchwork)
library(RColorBrewer)
library(ggsci)
library(car)
library(purrr)
library(broom)
library(ggpubr)
```

# Load the dataset

The first step is to load the dataset into R. You can load the dataset directly from the GitHub repository.

```{r, warning = F, message = F}
set.seed(2509)
## Load from GitHub repository

urlRemote  <- "https://raw.githubusercontent.com/brenosouzamarques/"
pathGithub <- "LANP_Esketamine_Adolescents/master/"
fileName   <- "Dataset.csv"

df <- paste0(urlRemote, pathGithub, fileName) %>% 
  read.csv()
```
\newpage

# Assess normality of outcomes

Q-Q Plots are used to assess the symmetry of dependent variables, that is, **MADRS total score** and **MADRS item 10** at baseline and 24-h post-infusion.

## MADRS total score at baseline and 24h post-treatment
```{r echo=TRUE, fig.align='center', fig.show='hold', message=FALSE, warning=FALSE, out.width='40%'}

ggqqplot(df$MADRS_Baseline)
ggqqplot(df$MADRS_Post)

```

## MADRS item 10 score at baseline and 24h post-treatment
```{r, fig.align='center', out.width= '40%', fig.show = 'hold', message = F, warning= F}
ggqqplot(df$MADRS_Suicide_Baseline)
ggqqplot(df$MADRS_Suicide_Post)

```

\newpage

# Descriptives

Total depressive symptoms and suicidal ideation scores are presented as means and standard deviations.

## MADRS total scores

```{r, warning = F, message = F}
madrs_total <- c(df$MADRS_Baseline, df$MADRS_Post)
time <- rep(c("baseline", "post"), each = 10)
data_madrs_total <- tibble(time, madrs_total)
kbl(data_madrs_total %>% group_by(time) %>% 
      summarise(
        Mean = mean(madrs_total),
        "Std dev" = sd(madrs_total),
        Maximum = max(madrs_total),
        Minimum = min(madrs_total),
        Median = median(madrs_total)),
    caption = "MADRS total score",
    booktabs = T) %>% kable_styling(latex_options = c("hold_position"))


```
|

## MADRS item 10 scores

```{r}
item_10_total <- c(df$MADRS_Suicide_Baseline, df$MADRS_Suicide_Post)
time <- rep(c("baseline", "post"), each = 10)
data_madrs_total <- tibble(time, item_10_total)
kbl(
  data_madrs_total %>% 
  group_by(time) %>% 
  summarise(
    Mean = mean(item_10_total),
    "Std dev" = sd(item_10_total),
    Maximum = max(item_10_total),
    Minimun = min(item_10_total),
    Median = median(item_10_total)),
  caption = "MADRS item 10 score",
  booktabs = T) %>% kable_styling(latex_options = c("hold_position"))
  
```

\newpage
# T-tests

We aimed to determine if a single 0.5mg/kg dose of esketamine can reduce depressive symptoms and suicidal ideation in adolescents. We test this hypothesis with a paired-samples t-test.

```{r}
### T-test - MADRS Total Scores


ttest_total_madrs <- t.test(x = df$MADRS_Baseline,
                            y = df$MADRS_Post,
                            paired = T,
                            alternative = "two.sided")


ttest_item10_madrs <- t.test(x = df$MADRS_Suicide_Baseline,
                             y = df$MADRS_Suicide_Post,
                             paired = T,
                             alternative = "two.sided")

table_madrs <- map_df(list(ttest_total_madrs, ttest_item10_madrs), tidy)

table_madrs$'Outcome' <- c("MADRS total score",
                           "MADRS item 10")

table_madrs <- table_madrs %>% rename(
  "Mean Difference" = estimate,
  "T statistic" = statistic,
  "DF" = parameter,
  "95% CI lower" = conf.high,
  "95% CI higher" = conf.low,
  "P-value" = p.value)

kbl(table_madrs[c(9, 4, 2, 3, 1, 5, 6)],
      caption = "T-tests", booktabs = T) %>% 
  kable_styling(latex_options = c("hold_position"))

```

\newpage

## Effect sizes

Effect sizes allow further interpretation of changes from baseline to 24 hours post-ketamine treatment.

```{r}
effectsize_madrs <- lsr::cohensD(df$MADRS_Baseline,
                                 df$MADRS_Post,
                                 method = "paired")

effectsize_item10 <- lsr::cohensD(df$MADRS_Suicide_Baseline,
                                  df$MADRS_Suicide_Post,
                                  method = "paired")

kbl(tribble(~Outcome, ~"Effect Size (Cohen's D)",
            "MADRS total score", effectsize_madrs,
            "MADRS item 10", effectsize_item10),
    caption = "Effect sizes",
    booktabs = T) %>% kable_styling(latex_options = c("hold_position"))
  
```

\newpage

# Bootstrapped 95% confidence intervals

## Create a bootstrap function

This functions creates a main element, the `data_dens` data frame. It stores the bootstrapped samples of means as they are calculated using the bias-corrected and accelerated (Bca) method. These are later used to plot density graphs that compare bootstrapped distributions of pre-and post-treatment symptoms.

```{r}
boot_funs <- function(df, f1, f2){
            a <- df[[f1]]
            b <- df[[f2]]
            boot_a <- boot(a,
                           function(x, j) mean(x[j], na.rm = T), #Boot baseline variable
                           R = 10000)
            boot_b <- boot(b,
                           function(x, j) mean(x[j], na.rm = T), #Boot 7 days variable
                           R = 10000)
            boot_ci_a <- boot.ci(boot.out = boot_a, #Calculate baseline CI
                                 conf = 0.95,
                                 type = "bca")
            boot_ci_b <- boot.ci(boot.out = boot_b, #Calculate post-treatment CI
                                 conf = 0.95,
                                 type = "bca")
            print(boot_ci_a) # Return baseline bootstrapped CI
            print(boot_ci_b) # Return post-treatment bootstrapped CI
            
            ## Code to extract bootstrapped sample means
            densitiy_a <- boot_a$t
            densitiy_b <- boot_b$t
            dens <- c(densitiy_a, densitiy_b)
            time <- rep(c("Baseline", "Post-Treatment"), each = 10000)
            data_dens <<- data.frame(dens, time)
            data_dens$time <- factor(data_dens$time)
}
```



\newpage
## Run bootstraps and create ggplot graph objects

### Bootstrap - MADRS total score
```{r}
### Bootstrap CI - MADRS total score
boot_funs(df, "MADRS_Baseline", "MADRS_Post")

plot_madrs <- ggplot(data_dens)+
            geom_density(aes(x = dens, fill = time),
                         color = "white",
                         alpha = 0.9)+
            labs(x = "Total MADRS scores")+
            scale_fill_brewer(name = "Time of assessment:",
                              palette = "Dark2")+
  theme_classic()+
  theme(text = element_text(family = "serif"),
        axis.text.x = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.text = element_text(size = 13),
        legend.title = element_text(size = 13)) ##ggplot2 graph
```
\newpage
### Bootstrap - MADRS item 10

```{r}
### Bootstrap CI - MADRS item 10

boot_funs(df, "MADRS_Suicide_Baseline", "MADRS_Suicide_Post")

plot_madrs10 <- ggplot(data_dens)+
            geom_density(aes(x = dens, fill = factor(time)),
                         color = "white",
                         alpha = 0.9)+
            labs(x = "Item 10 MADRS scores",
                 y = element_blank())+
            scale_fill_brewer(name = "Time of assessment:",
                              palette = "Dark2")+
  theme_classic()+
  theme(text = element_text(family = "serif"),
        axis.text.x = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.text = element_text(size = 13),
        legend.title = element_text(size = 13)) ##ggplot2 graph


```

\newpage
# Graphics

## Density plots of bootstrapped sample means

Creates an annotated plot using the `{ggplot2}` density plots for the bootstrapped sample means and the `{patchwork}` package.

```{r}
## Density plots of bootstrapped sample means

## Creates patchwork object
graficos <- plot_madrs + plot_madrs10 &
            theme(legend.position = "bottom",
                  text = element_text(family = "serif"),
                  title = element_text(size = 12),
                  legend.title = element_text(size = 12))

## Creates guided patchwork object
graficos_f <- graficos + plot_layout(guides = "collect")

## Edits and annotates patchwork object
plota <- graficos_f + plot_annotation(
    title = 'Supp. Fig. 1 - Bootstraped sample means for total and item 10 MADRS scores',
    subtitle = 'Density plots of results at baseline and 24 hours post-treatment')

## Saves plot in high resolution
ggsave("density_plot.png", plota, width=25, height=15, units="cm", dpi=450)
```

\newpage

## Main plots: pre-and post-treatment effects for depressive symptoms and suicidal ideation

This next part uses a long formatted dataset to plot pre-and post-infusion scores for each patient.

```{r}
## Create dataset for total MADRS scores
madrs_total <- c(df$MADRS_Baseline, df$MADRS_Post)
time <- rep(c("baseline", "post"), each = 10)
id <- rep(1:10, 2)
id <- factor(id)
madrs_data_plot <- tibble(id, time, madrs_total)

## plot total MADRS scores
madrs_plot <- madrs_data_plot %>% ggplot()+
            geom_boxplot(aes(x = time, y = madrs_total),
                         width = .3,
                         alpha = .3,
                         color = "grey")+
            scale_y_continuous(breaks = seq(from = 10,
                                            to = 50,
                                            by = 5))+
            scale_x_discrete(labels = c("Baseline", "24 hours \npost-treatment"))+
            geom_point(aes(x = time, y = madrs_total,
                           color = id,
                           group = id),
                       size = 1.5)+
            geom_line(aes(x = time, y = madrs_total,
                          color = id,
                          group = id,
                          alpha = .4),
                      size = 1.5)+
            scale_color_npg()+
            theme_classic()+
            guides(alpha = "none")+
            labs(x = "Time of assessment",
                 y = "Total MADRS scores",
                 color = "Patient ID")+
            theme(axis.title.x = element_text(size = 14,
                                              face = "bold"),
                  axis.text.x = element_text(size = 13),
                  axis.title.y = element_text(size = 15,
                                              face = "bold"),
                  axis.text.y = element_text(size = 13),
                  legend.position = "bottom",
                  legend.text = element_text(size = 13,
                                             face = "bold"))

## Create dataset for MADRS Item 10

si_pre <- df$MADRS_Suicide_Baseline
si_post <- df$MADRS_Suicide_Post

madrs_item10 <- c(si_pre, si_post)
time <- rep(c("baseline", "post"), each = 10)
id <- rep(1:10, 2)
id <- factor(id)
madrs_item_plot <- tibble(id, time, madrs_item10)

item10_plot <- madrs_item_plot %>% ggplot()+
            geom_boxplot(aes(x = time, y = madrs_item10),
                         width = .3,
                         alpha = .3,
                         color = "grey")+
            scale_y_continuous(breaks = c(0, 1, 2, 3, 4, 5, 6))+
            scale_x_discrete(labels = c("Baseline", "24 hours \npost-treatment"))+
            geom_point(aes(x = time, y = madrs_item10,
                           color = id,
                           group = id),
                       size = 1.5,
                       position = position_dodge(0.08))+
            geom_line(aes(x = time, y = madrs_item10,
                          color = id,
                          group = id,
                          alpha = .4),
                      size = 1.5,
                       position = position_dodge(0.08))+
            scale_color_npg()+
            theme_classic()+
            guides(alpha = "none")+
            labs(x = "Time of assessment",
                 y = "MADRS item 10 scores")+
            theme(axis.title.x = element_text(size = 14,
                                              face = "bold"),
                  axis.text.x = element_text(size = 13),
                  axis.title.y = element_text(size = 15,
                                              face = "bold"),
                  axis.text.y = element_text(size = 13),
                  legend.position = "bottom",
                  legend.text = element_text(size = 13,
                                             face = "bold"),
                  text = element_text(family = "serif"))

```

### Code to join plots and produce the final image using `{patchwork}`.

```{r, message = F, warning= F}
## Creates patchwork object
graficos_i10 <- (madrs_plot +
                   plot_layout(guides = "collect") &
                   theme(legend.position = "right",
                         text = element_text(family = "serif"))) +
                   (item10_plot +
                   theme(legend.position = "none"))

## Edits and annotates patchwork object
graficos_i10 <- graficos_i10 + plot_annotation(
    title = 'Total and item 10 MADRS scores at baseline and 24h post-treatment',
    subtitle = 'Scores are given for each patient') &
  theme(plot.title = element_text(size = 12),
        plot.subtitle = element_text(size = 11,
                                face = "italic"))

## Saves plot in high resolution
ggsave("MADRS_plot.png", graficos_i10, width=23, height=12, units="cm", dpi=650)
ggsave("MADRS_plot.tiff", graficos_i10, width=23, height=12, units="cm", dpi=450)
ggsave("MADRS_plot.pdf", graficos_i10, width=23, height=12, units="cm", dpi=1200)
```


  