---
title: "Esketamine in adolescents - pilot study"
author: "Breno Souza-Marques"
date: "24/04/2021"
output: html_document
df_print: paged
abstract: |
            This R Markdown file will exactly reproduce our data tidying, transformation, analysis and graphing processes in the R software. It requires that the dataset available in the Open Science Framework repository ATUALIZAR [https://osf.io/a3892/](https://osf.io/a3892/) is located in the same directory of this file. Please refer to the repository for a detailed description of each variable contained herein.
            All files in the repository, including this one, are licensed under [CC-BY-4.0](https://creativecommons.org/licenses/by/4.0/).
---
#### Required Packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=F, message=F}
# Required Packages
packages = c("tidyverse", "haven",
             "lsr", "rstatix", "exactRankTests", "boot", "skimr", "kableExtra",
             "patchwork", "RColorBrewer", "ggsci", "car")
package.check <- lapply(
            packages,
            FUN = function(x) {
                        if (!require(x, character.only = TRUE)) {
                                    install.packages(x)
                                    library(x, character.only = TRUE)
                        }
            }
)
```


### Load and manage dataset
To correctly load the dataset, place the "dataset.sav" file in the same directory of this R Markdown script.

```{r}
set.seed(252009)

## Read the database
df <- haven::read_sav("Dataset.sav")
df <- haven::zap_formats(df)

```


#### Perform Paired Samples Wilcoxon or T-tests for each outcome
```{r}
### Assess normality

## MADRS Total score
car::qqPlot(df$MADRS_pré)
car::qqPlot(df$MADRS_24h)

## MADRS Item 10 score
car::qqPlot(df$Item10_PréMADRS)
car::qqPlot(df$Item10_24hMADRS)

## Descriptives
madrs_total <- c(df$MADRS_pré, df$MADRS_24h)
time <- rep(c("baseline", "post"), each = 10)
data_madrs_total <- tibble(time, madrs_total)
data_madrs_total %>% group_by(time) %>% 
  summarise(
  mean = mean(madrs_total),
  sd = sd(madrs_total)
)


item_10_total <- c(df$Item10_PréMADRS, df$Item10_24hMADRS)
time <- rep(c("baseline", "post"), each = 10)
data_madrs_total <- tibble(time, item_10_total)
data_madrs_total %>% group_by(time) %>% 
  summarise(
  mean = mean(item_10_total),
  sd = sd(item_10_total)
)


### T-test - MADRS Total Scores

t.test(x = df$MADRS_pré,
       y = df$MADRS_24h,
       paired = T)
lsr::cohensD(df$MADRS_pré,
             df$MADRS_24h,
             method = "paired")

### T-test - MADRS Item 10 scores

t.test(x = df$Item10_PréMADRS,
       y = df$Item10_24hMADRS,
       paired = T)
lsr::cohensD(df$Item10_PréMADRS,
             df$Item10_24hMADRS,
             method = "paired")
```

### Create a bootstrap function
This functions creates a main element, the `data_dens` data frame. It stores the bootstrapped samples of means and medians for each neuropsychological test as they are calculated. These are later used to plot density graphs that compare bootstrapped sample means of pre-and-post assessment performance.
```{r}
boot_funs <- function(df, f1, f2){
            a <- df[[f1]]
            b <- df[[f2]]
            boot_a <- boot(a,
                           function(x, j) mean(x[j], na.rm = T), #Boot baseline variable
                           R = 10000)
            boot_b <- boot(b,
                           function(x, j) mean(x[j], na.rm = T), #Boot 7 days variable
                           R = 10000)
            boot_ci_a <- boot.ci(boot.out = boot_a, #Calculate baseline CI
                                 conf = 0.95,
                                 type = "bca")
            boot_ci_b <- boot.ci(boot.out = boot_b, #Calculate post-treatment CI
                                 conf = 0.95,
                                 type = "bca")
            print(boot_ci_a)
            print(boot_ci_b)
            
            ## Code to extract bootstrapped sample means
            densitiy_a <- boot_a$t
            densitiy_b <- boot_b$t
            dens <- c(densitiy_a, densitiy_b)
            time <- rep(c("Baseline", "Post-Treatment"), each = 10000)
            data_dens <<- data.frame(dens, time)
            data_dens$time <- factor(data_dens$time)
}
```

#### Run bootstraps
```{r}
### Bootstrap CI - MADRS total score
boot_funs(df, "MADRS_pré", "MADRS_24h")

plot_madrs <- ggplot(data_dens)+
            geom_density(aes(x = dens, fill = time),
                         color = "white",
                         alpha = 0.9)+
            labs(x = "Total MADRS scores")+
            scale_fill_brewer(name = "Time of assessment:",
                              palette = "Dark2")+
  theme_classic()+
  theme(text = element_text(family = "serif"),
        axis.text.x = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.text = element_text(size = 13),
        legend.title = element_text(size = 13))



### Bootstrap CI - MADRS item 10

boot_funs(df, "Item10_PréMADRS", "Item10_24hMADRS")

plot_madrs10 <- ggplot(data_dens)+
            geom_density(aes(x = dens, fill = factor(time)),
                         color = "white",
                         alpha = 0.9)+
            labs(x = "Item 10 MADRS scores",
                 y = element_blank())+
            scale_fill_brewer(name = "Time of assessment:",
                              palette = "Dark2")+
  theme_classic()+
  theme(text = element_text(family = "serif"),
        axis.text.x = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.text = element_text(size = 13),
        legend.title = element_text(size = 13))


```

### Graphics

This first part creates an annotated plot using the `ggplot2` density plots for the bootstrapped sample means.
```{r}
## Density plots of bootstrapped sample means

## Creates patchwork object
graficos <- plot_madrs + plot_madrs10 &
            theme(legend.position = "bottom",
                  text = element_text(family = "serif"),
                  title = element_text(size = 12),
                  legend.title = element_text(size = 12))

## Creates guided patchwork object
graficos_f <- graficos + plot_layout(guides = "collect")

## Edits and annotates patchwork object
plota <- graficos_f + plot_annotation(
            title = 'Supp. Fig. 1 - Bootstraped sample means for total and item 10 MADRS scores',
            subtitle = 'Density plots of results at baseline and 24 hours post-treatment')
plota


## Saves plot in high resolution
ggsave("density_plot.png", plota, width=25, height=15, units="cm", dpi=450)
```

This next part uses a long formatted dataset to compare pre-and post-infusion results for each patient.
```{r}
## Create dataset for total MADRS scores
madrs_total <- c(df$MADRS_pré, df$MADRS_24h)
time <- rep(c("baseline", "post"), each = 10)
id <- rep(1:10, 2)
id <- factor(id)
madrs_data_plot <- tibble(id, time, madrs_total)

## plot total MADRS scores
madrs_plot <- madrs_data_plot %>% ggplot()+
            geom_boxplot(aes(x = time, y = madrs_total),
                         width = .3,
                         alpha = .3,
                         color = "grey")+
            scale_y_continuous(breaks = seq(from = 10,
                                            to = 50,
                                            by = 5))+
            scale_x_discrete(labels = c("Baseline", "24 hours post-infusion"))+
            geom_point(aes(x = time, y = madrs_total,
                           color = id,
                           group = id),
                       size = 1.5)+
            geom_line(aes(x = time, y = madrs_total,
                          color = id,
                          group = id,
                          alpha = .4),
                      size = 1.5)+
            scale_color_npg()+
            theme_classic()+
            labs(x = "Time of assessment",
                 y = "Total MADRS scores")+
            theme(axis.title.x = element_text(size = 14),
                  axis.text.x = element_text(size = 13),
                  axis.title.y = element_text(size = 15),
                  axis.text.y = element_text(size = 13),
                  legend.position = "none")

## Create dataset for MADRS Item 10

si_pre <- df$Item10_PréMADRS
si_post <- df$Item10_24hMADRS

madrs_item10 <- c(si_pre, si_post)
time <- rep(c("baseline", "post"), each = 10)
id <- rep(1:10, 2)
id <- factor(id)
madrs_item_plot <- tibble(id, time, madrs_item10)

item10_plot <- madrs_item_plot %>% ggplot()+
            geom_boxplot(aes(x = time, y = madrs_item10),
                         width = .3,
                         alpha = .3,
                         color = "grey")+
            scale_y_continuous(breaks = seq(from = 0,
                                            to = 6,
                                            by = 1))+
            scale_x_discrete(labels = c("Baseline", "24 hours post-infusion"))+
            geom_point(aes(x = time, y = madrs_item10,
                           color = id,
                           group = id),
                       size = 1.5,
                       position = position_dodge(0.08))+
            geom_line(aes(x = time, y = madrs_item10,
                          color = id,
                          group = id,
                          alpha = .4),
                      size = 1.5,
                       position = position_dodge(0.08))+
            scale_color_npg()+
            theme_classic()+
            labs(x = "Time of assessment",
                 y = "MADRS item 10 scores")+
            theme(axis.title.x = element_text(size = 14),
                  axis.text.x = element_text(size = 13),
                  axis.title.y = element_text(size = 15),
                  axis.text.y = element_text(size = 13),
                  legend.position = "none",
                  text = element_text(family = "serif"))

## Creates patchwork object
graficos_i10 <- madrs_plot + item10_plot &
            theme(text = element_text(family = "serif"),
                  title = element_text(size = 16),
                  legend.title = element_text(size = 14))

## Edits and annotates patchwork object
graficos_i10 <- graficos_i10 + plot_annotation(
            title = 'Total and item 10 MADRS scores at baseline and 24h post-treatment',
            subtitle = 'Scores are given for each patient')
graficos_i10


## Saves plot in high resolution
ggsave("MADRS_plot.png", graficos_i10, width=25, height=15, units="cm", dpi=500)

```

